Bacteriophage infection drives loss of β-lactam resistance in methicillin-resistant Staphylococcus aureus

Version 1

Abstract

Abstract

Bacteriophage (phage) therapy has been proposed as a means to combat drug-resistant bacterial pathogens. Infection by phage can select for mutations in bacterial populations that confer resistance against phage infection. However, resistance against phage can yield evolutionary trade-offs of biomedical use. Here we report the discovery of staphylococcal phages that cause different strains of methicillin-resistant Staphylococcus aureus (MRSA) to become sensitized to β-lactams, a class of antibiotics against which MRSA is typically highly resistant. MRSA cells that survive infection by these phages display significant reductions in minimal inhibitory concentration against different β-lactams compared to uninfected bacteria. Phage-treated MRSA further exhibited attenuated virulence phenotypes in the form of reduced hemolysis and clumping. Sequencing analysis revealed that the different MRSA strains evolved unique genetic profiles during infection. These results suggest complex evolutionary trajectories in MRSA during phage predation and open up new possibilities to reduce drug resistance and virulence in MRSA infections.

Staphylococcus aureus is a Gram-positive bacterium found on the skin and mucosa of humans and animals. It causes a wide range of human diseases, from endocarditis to pneumonia.1 Methicillin-resistant S. aureus (MRSA) is a form of S. aureus that displays high levels of drug resistance towards β-lactam antibiotics, such as penicillins and cephalosporins. β-lactams act by inhibiting the activity of transpeptidase enzymes during peptidoglycan synthesis in bacterial cell walls.2 β-lactams are one of the most commonly prescribed drug classes, and many are designated as “Critically Important” antimicrobials by the World Health Organization.3

A chief mediator of β-lactams resistance in MRSA is the SCCmec cassette, a mobile genetic element that carries the resistance gene mecA. MecA encodes for the penicillin-binding protein 2A (PBP2a), a transpeptidase that has a low affinity for β-lactams.4 This lower affinity permits PBP2a to participate in peptidoglycan synthesis even in the presence of β-lactams, ultimately resulting in cell survival. In addition to mecA, numerous MRSA strains carry β-lactamases, such as BlaZ, that degrade β-lactams, thus further contributing to drug resistance. MRSA infections pose a considerable public health risk as they are notoriously difficult to treat and are widespread in communities and hospital settings. In 2019, MRSA alone accounted for more than 100,000 deaths attributable to drug-resistant infections worldwide.5 Developing novel solutions to combat MRSA is a major focus in academia and industry.

Bacteriophages (phages) are viruses that infect and kill bacteria, posing one of the greatest existential threats to bacterial communities. Some estimates suggest that 40% of all bacterial mortality worldwide is caused by phage predation.6 Due to this lethality, phage therapy has been proposed as an attractive alternative for combating drug-resistant bacterial infections. In phage therapy, lytic phages are administered to kill the bacterial pathogen(s) causing an infection. Phages offer certain advantages over traditional antibiotics: they are highly specific for their hosts by reducing off-target killing; they self-amplify and evolve, enabling the rapid generation of new phage variants with improved activities; and they are generally regarded as safe, based on extremely rare cases of toxicity in animals and patients.7 Indeed, against staphylococcal infections, over a dozen promising case studies and clinical trials have been reported in the past decade.8 Despite these advances, routine use of phage therapy is still met with challenges.

Chief among these challenges is the inevitable rise of phage resistance as phage predation exerts a strong selective pressure on bacterial populations. According to one analysis that focused on phage therapy outcomes, resistance against phage evolved in 75 percent of human clinical cases in which the evolution of resistance was monitored.9 Mutations represent a chief pathway by which bacteria evolve resistance against phage. To date, the best characterized phage resistance mutations strains involve alterations of the cell surface that affect phage attachment. For most lytic phages, the first step of their infection involves attachment of the phage particle to a receptor molecule (or multiple molecules) on the cell surface; these receptors are often proteins or sugar moieties, which are recognized by phage proteins.8,10,11 Mutations in these receptor molecules can prevent phage binding, leading to resistance. For example, in Escherichia coli, mutations in the cell-wall protein LamB confer resistance against lambda phage infection, while in S. aureus, mutations that modify the glycopolymers in the cell wall, such as wall teichoic acid (WTA), have been shown to limit the infection by phages.11,12 Complicating the picture, studies have revealed that a plethora of additional host mechanisms, including dedicated anti-phage defense systems, can impact the evolution of resistance against phage.13,14 These problems highlight the importance of developing phage treatment strategies that minimize or capitalize on the evolution of phage resistance.15

A unique aspect of phage therapy is the possibility to exploit evolutionary trade-offs to combat resistant pathogens. A genetic trade-off is defined as an evolved trait that confers a fitness advantage against a particular selective pressure at the expense of reduced fitness against an unselected pressure. Across many different species of bacteria, such fitness trade-offs are known to occur between phage resistance and antibiotic resistance (Figure 1A). For example, selection by a lytic phage PYOSa on the methicillin-sensitive S. aureus Newman selected for mutations in the femA aminoacyltransferase, leading to increased resistance against phage infection but reduced resistance towards antibiotics.16 These studies raise the question of whether such phage/antibiotic resistance trade-offs can be induced in actual MRSA strains, which would have significant therapeutic implications. Employing phages to drive the evolution of such beneficial trade-offs could serve as a means to reduce drug resistance. However, very little is known about how frequently these trade-offs occur in different bacterial species, or through what molecular mechanisms they work.

Infection by bacteriophage ΦStaph1N drives the loss of β-lactam resistance in MRSA.

A. Schematic of the experimental setup. Drug-resistant (AbxR), phage-sensitive (PhageS) bacterial cultures are infected with phage. The population of infected cells is passaged and allowed to recover. The surviving cell population is resistant to phage infection (PhageR) but has evolved sensitivity to antibiotics (AbxS). B. ΦStaph1N infects MRSA strains MRSA252, MW2, and LAC (left panel). Following infection with ΦStaph1N, evolved cultures of the three MRSA strains are resistant to ΦStaph1N (right panel). C. ΦStaph1N-treated, evolved MRSA strains show significant loss of resistance against oxacillin, compared to the parental strains. Loss of resistance is indicated by the area of bacterial clearance surrounding the antibiotic resistance strip. D. ΦStaph1N treatment causes loss resistance against different β-lactams. Plotted are the fold reductions of minimal inhibitory concentration (MIC) between treated and mock-treated cells. OXI = oxacillin; CEF = cefazolin; AMX = amoxicillin; AMX+CA = amoxicillin & clavulanic acid; VANC = vancomycin.

Here we investigate how infection by staphylococcal phages causes three different MRSA strains to evolve sensitivity to different types of β-lactam antibiotics. Interestingly, despite being treated by the same phage, the loss of resistance phenotype is mediated by mutations distinct to each MRSA strain. In addition to resistance, phage infection also causes a reduction in hemolysis and clumping, which are key virulence phenotypes in MRSA. Finally, we isolated a mutant version of the phage that displays enhanced killing, improved resistance reduction, and reduced virulence phenotypes across all three MRSA strains.

Results

Identification of ΦStaph1N with activity against multiple MRSA strains

We subjected three MRSA strains – MRSA252 (USA200), MW2 (USA400), and LAC (USA300) – to a panel of staphylococcal phages using plaquing assays (Table S1, Figure S1). All three MRSA strains are pathogenic isolates that have been implicated in human infection and disease and have been used as representative examples for studying MRSA.17–19 Of the phages tested, one phage, ΦStaph1N, was able to form distinct plaques on all three MRSA strains (Figure 1B). ΦStaph1N belongs to the myovirus family of lytic staphylococcal phages, with a genome size of ∼138 kB.20,21

Despite its ability to infect all three MRSA strains, ΦStaph1N infection was often unable to effectively lyse MRSA cultures. For example, both MW2 and LAC displayed incomplete lysis in liquid culture at multiplicities of infection (MOIs) of 0.1 or lower (Figure S2). At an MOI of 0.1, we observed that all three MRSA strains lysed within 2-3 hours following phage infection but recovered high cell density after passaging one percent of the culture into fresh media. We therefore hypothesized that predation by ΦStaph1N selected for mutations conferred resistance against ΦStaph1N infection, allowing the culture to regrow. Indeed, ΦStaph1N was unable to form plaques on recovered MRSA cultures that survived in the initial ΦStaph1N infection (Figure 1B).

Next, we asked whether phage resistance against ΦStaph1N infection could cause a trade-off in drug resistance in MRSA. Since both phage and β-lactams interface with the bacterial cell wall, we hypothesized that phage resistance mutations could impact β-lactam activity. Parental strains of MRSA252, MW2, and LAC displayed higher minimal inhibitory concentrations (MICs) of ≥48 µg/mL against the β-lactams oxacillin (OXA), cefazolin (CEF), amoxicillin (AMX), and amoxicillin & clavulanic acid (AMX+CA), visually indicated by their ability to form lawns surrounding antibiotic strips (Figure 1C, Table S2). The MRSA strains were sensitive to vancomycin (VANC; MICs = 1.5 µg/mL), which inhibits cell wall synthesis through a different mechanism than β-lactams.

By contrast, all strains that survived ΦStaph1N infection displayed a reduction in resistance against OXA, CEF, and AMX+CA, with fold reductions in MIC between 10 and 1000-fold (Figure 1D). No changes in MIC were observed with VANC. These observations suggest that the loss of resistance might be specific to the mechanism of action of β-lactams. Furthermore, phage-treated strains remained resistant to AMX alone, likely caused by the activity of β-lactamases present in the three MRSA strains. Together, these results suggest that predation by ΦStaph1N causes three different MRSA strains to lose their resistance towards β-lactams.

Next, we asked how infection at lower MOIs with ΦStaph1N affected β-lactam resistance in MRSA. We infected the three MRSA strains with ΦStaph1N at MOIs ranging from 0.01 to 10-5, isolated the surviving MRSA cells, and tested the survivors for their MIC against oxacillin; each condition was tested in three independent replicates (Table S2). For MRSA252, we observed a ∼3-order of magnitude fold reduction of MIC in the surviving cells at an MOI of 10-5. With MW2, the reduction of MIC markedly decreased with lower phage levels, showing on average no reduction in MIC at MOIs of 10-3 or lower (Table S2). For LAC, two of the replicates displayed an order of magnitude reduction of MIC at an MOI of 10-4, while one replicate did show a strong decrease in MIC at the same MOI. These results suggest that for MRSA252, MOIs as low as 10-5 of ΦStaph1N infection can still drive the loss of resistance, while for MW2 and LAC, higher MOIs of phage are needed to ensure the same outcome of reduced β-lactam resistance.

We then asked whether the reduction in β-lactam can be achieved with a phage from a different taxonomical group. The MRSA LAC strain is sensitive to infection by <ΙNM1ψ6, a lytic version of the temperate phage <ΙNM1, displaying plaquing comparable to ΦStaph1N (Table S1, Figure S1). <ΙNM1ψ6 belongs to the siphovirus family and was derived from the S. aureus Newman strain.22–24 We therefore examined whether infection with <ΙNM1ψ6 could also reduce β-lactam resistance in LAC. We infected LAC with <ΙNM1ψ6 at an MOI of 0.1 and measured the MIC against oxacillin of the surviving cells. Here, recovered LAC cultures did not show a reduction in MICs (Figure S3). This suggests that <ΙNM1ψ6 is unable to reduce resistance in susceptible MRSA hosts, raising the possibility that different phages vary in their ability to change β-lactam resistance in MRSA.

Discovery of mutant in ΦStaph1N with enhanced activity against MRSA

A key advantage of phage therapy over conventional pharmaceutical antibiotics is that phages themselves can evolve. For ΦStaph1N, we noticed that while the phage was able to plaque all three MRSA strains, its plaque-forming efficiency was reduced on the MW2 and LAC strains (Figure 1B, Figure S1). ΦStaph1N plaques on MW2 and LAC strains were hazy and the overall efficiency of plaquing was approximately two orders of magnitude less than that with MRSA252. Unexpectedly, we consistently observed smaller, clear plaques arising in the larger, hazy plaques of LAC (Figure 2A). This led us to hypothesize that these clear plaques are caused by a mutant ΦStaph1N that plaques more efficiently on LAC. We therefore isolated phage clones from these single plaques and tested for their activity against MRSA. This mutant phage, which we called Evo2, is able to plaque on LAC and MW2 strains with higher efficiency, displaying comparable plaquing to MRSA252 (Figure 2A, Figure S1). In growth curve experiments, we further observed that Evo2 lyses MRSA cultures at lower MOIs compared to ΦStaph1N (Figure S2). Notably, Evo2 exhibits lytic activity against MW2 and LAC even at an MOI of 0.0001, a concentration at which ΦStaph1N does not show any detectable activity against the two strains.

Evo2 is a variant of ΦStaph1N with higher activity against MRSA.

A. Single Evo2 plaques appear inside hazy ΦStaph1N plaques on LAC (Left-most panel). Compared to ΦStaph1N, Evo2 shows comparable infectivity towards MRSA252 but improved infectivity towards MW2 and LAC (Right-most panels). B. Similar to ΦStaph1N, Evo2 infection reduces β-lactam resistance in MRSA.

We sequenced the genome of Evo2 to determine the genetic mechanism driving this enhanced activity. We observed a single point mutation in ORF141 that induces a premature stop codon (Figure S4). Sequence analysis with HHpred predicts ORF141 to be a putative DNA binding protein with an HTH motif (PDB: 2LVS, E-value: 2.5e-9). We speculate that this protein is a transcriptional regulator that when inactivated by a nonsense mutation, increases ΦStaph1N infectivity. Future studies will center on determining the mechanism of this mutation and why Evo2 only evolved in the LAC strain. Given Evo2’s enhanced activity against MRSA, we then asked how predation by Evo2 impacted β-lactam resistance in our three MRSA strains. We infected MRSA252, MW2, and LAC with Evo2 at an MOI of 0.1 and measured the MICs against β-lactams after 48 hours of passaging. Broadly, similar to the ΦStaph1N infection, infection by Evo2 reduced the MICs of the three MRSA strains against OXA, CEF, and AMX+CA, while MICs against AMX alone and VAN did not change significantly (Figure 2B).

We then tested how different MOIs of Evo2 impacted β-lactam resistance in MRSA (Table S2). As with ΦStaph1N, we infected the three MRSA strains with Evo2 at varying MOIs from 10-2 to 10-5 and measured the MIC against oxacillin in the evolved MRSA; each condition was tested in three independent replicates. For most of the MRSA252 cultures, we did not see a significant difference in the fold reduction of MIC between ΦStaph1N and Evo2. Across all three strains, we observed that replicate cultures across different MOIs were unable to recover growth following Evo2 infection (Table S2). However, cultures of MW2 and LAC that did regrow displayed a loss of oxacillin resistance. Thus, overall Evo2 displays a higher infectivity against the three MRSA strains and a greater potency in reducing β-lactam resistance.

Genomic mutations in MRSA strains following phage infection

Following these MIC tests, we hypothesized that the phage-treated MRSA stains evolved mutations that increased their sensitivity to β-lactams. We sequenced the genomes of three clonal isolates (A-C) from each MRSA strain that underwent ΦStaph1N, Evo2, or a mock infection. We observed that each MRSA strain evolved distinct mutation profiles (Figure 3A). Irrespective of the strain and phage treatment, most mutations were predicted to result in substitutions in the protein product, followed by truncations (Figure S5). Cluster of Orthologous Genes (COG) variants associated with transcription, cell wall/membrane/envelope biogenesis, coenzyme transport and metabolism, and defense mechanisms, were the most commonly found categories. Mutations in annotated genes that appeared at least twice across the clonal replicates are summarized in Table 1. Information on all detected genetic variants is listed in Supplementary Table 1.

Phage infection of MRSA strains produces distinct mutational profiles.

A. Coding sequences (CDS) with mutations from the three MRSA strains following phage treatment or mock treatment. For each strain, three isolates were sequenced and their mutations identified. Mutations are color-coded based on the number of occurrences among the three replicates. B. Categories of genes with mutations that arose in each MRSA strain and treatment condition.

In all MRSA strains, we found mutations that could account for the loss in β-lactam resistance. In MRSA252, both ΦStaph1N and Evo2 infection were selected for frameshift or nonsense mutations in the femA gene that would inactive the protein product. FemA is required for the synthesis of the pentaglycine branch on S. aureus Lipid II, the peptidoglycan precursor (Table 1). Deletions of femA have been shown to increase susceptibility to β-lactams even when PBP2a (encoded by mecA) is expressed.37,38 These results thus provide a potential mechanism for how some MRSA252 cells lose β-lactam resistance after phage selection. At the same time, we noted the presence of other uncharacterized mutations in phage-treated MRSA252. For example, clone A of ΦStaph1N-treated cells carried 2 mutations: a frameshift in femA and a substitution mutation in an uncharacterized protein; meanwhile, clone B displayed a substitution mutation in pfkA, a predicted ATP-dependent 6-phosphofructokinase and mutation in an intergenic region; clone C showed a substitution mutation in a putative transport protein, called yueF (Figure 3B, Supplementary Table 1). The role of these mutations in mediating phage resistance or β-lactam sensitivity, if any, remains unknown.

In MW2, we found mutations in two transcriptional regulators, mgrA and sarA (Figure 3A, Table 1). Both mgrA and sarA belong to the family of MarR (multiple antibiotic resistance regulator)/SarA (staphylococcal accessory regulator A) proteins, which regulate drug resistance and virulence in S. aureus.25–27 In ΦStaph1N-treated MW2, only mgrA was mutated, while in Evo2-treated MW2, clones also showed nonsense mutations in sarA. We also found mutations in metK and ytqA which are both predicted to be associated with S-adenosylmethione: metK synthesizes SAM, while ytqA belongs to the radical SAM enzyme family and is predicted to be involved in tRNA modification.35,40 Phage-treated MW2 also displayed mutations in other genes, including tcaB and murE. Notably, each clonal replicate had multiple mutations in the genome, while by contrast untreated MW2 cells only displayed a deletion in an intergenic region that is not present in any of the phage-treated samples.

In the LAC strain, we observed a third, distinct mutational pattern (Figure 3A). Of note, we found nonsense mutations in arlR, which is part of the arlRS two-component signaling system (Table 1). The activity of ArlRS has been implicated in S. aureus virulence, pathogenicity, and oxacillin resistance.30,31 Further, we observed substitution mutations in spoVG, which is a transcription factor regulating the expression genes involved in a variety of functions, including cell wall metabolism (Table 1).32 Indeed, SpoVG activates the expression of femA.33 Studies have shown that SpoVG modulates β-lactam antibiotic resistance by modulating cell wall synthesis.33 Similar to the other two MRSA strains, phage-treated clones of LAC showed multiple mutations in their genomes, although their association with β-lactam sensitivity remains unclear. Finally, we observed mutations in prsA and bioA that appeared in both the mock and phage treatment conditions, suggesting that these mutations do not arise due to phage selection.

Phage-treated MRSA strains display reduced virulence phenotypes

S. aureus strains encode a wide variety of virulence factors that facilitate bacterial infection, and previous studies have used MW2 and LAC strains as models of MRSA virulence.27,28 In MW2 and LAC, we observed the evolution of nonsense mutations in known virulence regulators, such as mgrA, arlR, and sarA. We therefore hypothesized that phage-treated MRSA cells could also display altered virulence phenotypes in addition to reduced β-lactam resistance. We first tested the ability of MRSA strains that survived phage predation for their ability to form biofilms in a Crystal Violet assay. We found that Evo2 infection of MRSA252 resulted in a significant reduction in Crystal Violet absorption compared to the parental strain. However, we show no significant difference in Crystal Violet absorption between parental, mock- and phage-treated MW2 and LAC strains (Figure S6).

Next, we tested whether phage infection could affect the hemolysis of rabbit blood cells. Hemolysis is mediated by the secretion of toxins, such as alpha toxins, and plays an important role in MRSA infection.43 Expression of these toxins is regulated by virulence pathways that comprise numerous transcription factors, including mgrA, arlR, and sarA. Parental MW2 and LAC colonies lysed rabbit blood cells on blood agar plates, producing distinct halos of clearance around the bacterial cells. As expected, for untreated LAC, the total area of hemolysis was on average 3-fold larger than that of untreated MW2 (∼210 mm2 vs ∼70 mm2, respectively); with MRSA252, by contrast, lysis was not detected (Figure 4A). Following treatment with ΦStaph1N, we observed that MW2 and LAC displayed a reduced area of hemolysis by 4 to 5-fold. With Evo2-treated cells, we found that in MW2 the fold reduction was comparable to that of ΦStaph1N-treated cells. However, for Evo2-treated LAC, loss of hemolysis was even more pronounced, with two of the replicates showing no detectable hemolysis.

Phage treatment of MRSA results in attenuated virulence phenotypes.

A. MW2 and LAC strains display hemolytic activity on rabbit blood agar plates, while MRSA252 does not. B. Phage-treated MW2 and LAC strains display reduced hemolysis compared to uninfected cells. C. Surviving cultures of MW2 and LAC treated with either ΦStaph1N (blue) or Evo2 (red) show reduced clumping rates compared to mock untreated cells (teal).

S. aureus is also able to bind to fibrinogen, resulting in the agglutination, or clumping, of bacteria. Clumping is thought to have several functions in the context of staphylococcal infections. Clumps are likely to be more resistant to clearance by the immune system, partly because they may be too large to be phagocytosed by neutrophils.28 Similar to the expression of hemolysins, previous work has shown that clumping in MW2 and LAC is regulated by a network of virulence regulators. For MW2 and LAC, we observed that phage-treated cells displayed less clumping than the mock-treated or parental strain. For MW2, ΦStaph1N infection resulted in a modest reduction, while Evo2 infection resulted in a reduction of approximately 3-fold (Figure 4B). In LAC, we found that both ΦStaph1N and Evo2 treatment resulted in comparable reductions of clumping in surviving cells.

Discussion

Here we report the discovery that infection by the myovirus ΦStaph1N and its derivative Evo2 can drive MRSA populations to evolve favorable genetic trade-offs between phage and β-lactam resistance. MRSA strains that survived phage infection were resistant to phage; however, they displayed up to a 1000-fold decrease in their MICs against β-lactam antibiotics. Sequencing revealed that each of the three MRSA strains tested evolved distinct mutational profiles following phage infection. Of these mutations, a significant portion of the affected genes are involved in transcriptional regulation, antibiotic resistance, and cell maintenance. Concurring with the sequencing data, phage-treated MRSA strains displayed attenuated virulence phenotypes in the form of reduced hemolysis and cellular agglutination.

Our system highlights the multitude of evolutionary paths that different bacterial strains can undertake to evolve resistance against phage. Most staphylococcal phages are thought to adsorb to S. aureus by recognizing wall teichoic acid (WTA). Prior work has revealed that phage resistance in S. aureus can arise from the disruption of genes involved in the synthesis and modification of WTA, such as tagO or tarM.44 Some MRSA strains also alter cell wall glycosylation through dedicated genes encoded on prophages.11 Additionally, the presence of cell wall-bound Protein A has been shown to decrease phage absorption, likely by masking WTA.45 However, our results suggest that the evolution of phage resistance can potentially impact networks of genes in the different MRSA strains, including metabolic genes and transcription factors. For example, both mgrA and arlRS regulate the expression of proteins associated with the cell wall, such as the Giant Staphylococcal Surface Protein (GSSP).28 We hypothesize that mutations in these regulators might perturb the composition of proteins in the cell wall, which in turn can disrupt phage binding or prevent disrupt injection of the phage nucleic acid into the cytoplasm. Further studies are needed to dissect the effect of these gene networks on phage resistance and β-lactam resistance.

Our study also shows some of the challenges associated with determining phage/antibiotic resistance trade-offs. Not only do different bacterial strains evolve distinct mutations, but individual clones can accumulate multiple mutations in their genome. Such complex dynamics have been observed in other coevolving bacteria-phage systems before and can serve as useful models to study evolution. However, they make it difficult to predict broader phenotypic outcomes in the bacteria following phage infection. For example, phage resistance has also been shown to confer so-called “trade-ups,” where a non-selected trait is enhanced (e.g. cross-resistance between phage and antibiotic). In our study, we cannot rule out that our phage-treated MRSA has evolved (or will evolve) trade-ups. Although further research is required to test this possibility, the potential clinical benefit of reduced resistance and virulence in MRSA may outweigh the risks associated with trade-ups.

Despite evolving distinct mutational profiles, the three MRSA strains nonetheless displayed a convergence of phenotypes in the form of enhanced phage resistance, reduced β-lactam resistance, and attenuated virulence. We posit that this convergence is caused by the involvement of the cell wall in all three phenotypic outcomes. Phages must interface with the cell wall to adsorb, β-lactams target proteins associated with the maintenance of the cell wall, and many S. aureus virulence factors are embedded within the cell wall or need to get secreted through it. The integrity and composition of the cell wall can be determined by numerous genes, ranging from single proteins involved in cell wall synthesis, such as femA, to transcriptional regulators that control the expression of cell wall-associated proteins, such as mgrA. Thus, the mutational patterns observed in each MRSA strain might reflect unique genetic solutions that enable the bacteria to adapt to the phage predation, while also maximizing the fitness for that particular strain.

From a biomedical perspective, exploiting phage/antibiotic resistance trade-offs have been proposed as a means to combat resistance in bacterial pathogens.15 Not only could phage treatments reduce the bacterial load of an infection, but also potentially resensitize the bacterial population to antibiotics that it was previously resistant against. For example, in our study, we observed that the MIC of phage-treated MRSA against oxacillin dropped below 2 µg/mL, which is designated by the FDA as the oxacillin MIC susceptible breakpoint for S. aureus. We speculate that infection by ΦStaph1N, Evo2, or an equivalent phage can facilitate the re-deployment of β-lactams in the treatment against MRSA infections. Additionally, phage-treated MRSA might also lose virulence and become more susceptible to immune clearance, which can further improve treatment outcomes. Future studies will need to examine this hypothesis in animal infection models.

Supporting information

eLife Assessment

The manuscript explores how bacterial evolution in the presence of lytic phages modulates b-lactams resistance and virulence properties in methicillin-resistant Staphylococcus aureus (MRSA). The work is valuable as it identifies underlying mutations that may confer sensitivity to b-lactams and alter virulence properties. While the findings are generally convincing, additional experiments linking how particular mutations regulate phenotypic changes are required to improve the work mechanistically.

Reviewer #1 (Public review):

Summary:

These authors have asked how lytic phage predation impacts antibiotic resistance and virulence phenotypes in methicillin-resistant Staphylococcus aureus (MRSA). They report that staphylococcal phages cause MRSA strains to become sensitized to b-lactams and to display reduced virulence. Moreover, they identify mutations in a set of genes required for phage infection that may impact antibiotic resistance and virulence phenotypes.

Strengths:

Phage-mediated re-sensitization to antibiotics has been reported previously but the underlying mutational analyses have not been described. These studies suggest that phages and antibiotics may target similar pathways in bacteria.

Weaknesses:

One limitation is the lack of mechanistic investigations linking particular mutations to the phenotypes reported here. This limits the impact of the work.

Another limitation of this work is the use of lab strains and a single pair of phages. However, while incorporation of clinical isolates would increase the translational relevance of this work it is unlikely to change the conclusions.

Reviewer #2 (Public review):

Summary:

The work presented in the manuscript by Tran et al deals with bacterial evolution in the presence of bacteriophage. Here, the authors have taken three methicillin-resistant S. aureus strains that are also resistant to beta-lactams. Eventually, upon being exposed to phage, these strains develop beta-lactam sensitivity. Besides this, the strains also show other changes in their phenotype such as reduced binding to fibrinogen and hemolysis.

Strengths:

The experiments carried out are convincing to suggest such in vitro development of sensitivity to the antibiotics. Authors were also able to "evolve" phage in a similar fashion thus showing enhanced virulence against the bacterium. In the end, authors carry out DNA sequencing of both evolved bacteria and phage and show mutations occurring in various genes. Overall, the experiments that have been carried out are convincing.

Weaknesses:

Although more experiments are not needed, additional experiments could add more information. For example, the phage gene showing the HTH motif could be reintroduced in the bacterial genome and such a strain can then be assayed with wildtype phage infection to see enhanced virulence as suggested. At least one such experiment proves the discoveries regarding the identification of mutations and their outcome. Secondly, I also feel that authors looked for beta-lactam sensitivity and they found it. I am sure that if they look for rifampicin resistance in these strains, they will find that too. In this case, I cannot say that the evolution was directed to beta-lactam sensitivity; this is perhaps just one trait that was observed. This is the only weakness I find in the work. Nevertheless, I find the experiments convincing enough; more experiments only add value to the work.

Author response:

The following is the authors’ response to the original reviews

We sincerely thank the reviewers for their thoughtful review and feedback. We believe that our work will provide valuable insights into how MRSA evolves under bacteriophage predation and stimulate efforts to use genetic trade-offs to combat drug resistance. We have substantially revised the paper and performed several additional experiments to address the reviewers' questions and concerns.

Summary:

(1) Testing for genetic trade-offs in additional S. aureus strains

We obtained 30 clinical isolates of the S. aureus USA300 strain that were isolated between 2008 and 2011 (see Table S1). We first tested the FStaph1N, Evo2, and FNM1g6 phages against this expanded strain panel and found that Evo2 showed strong activity against all 30 strains (Table S4). We tested whether Evo2 infection could elicit trade-offs in b-lactam resistance for a subset of these strains. We found that Evo2 infection caused a ~10-100-fold reduction in their MIC against oxacillin. This data is now incorporated into a revised Figure 2 in panel C.

(2) Testing additional staphylococcal phages

We isolated from the environment a phage called SATA8505. Similar to FStaph1N and Evo2, SATA8505 belongs to the Kayvirus genus and infects the MRSA strains MRSA252, MW2, and LAC. Phage-resistant MRSA recovered following SATA8505 infection also showed a strong reduction in oxacillin resistance (Figure S5). Furthermore, we confirmed that resistance against FNM1g6, which belongs to the Dubowvirus genes, does not elicit tradeoffs in b-lactam resistance (Figure S4). Sequencing analysis of FNM1g6 - resistant LAC strains showed a different mutation fmhC, which was not observed with the FStaph1N and Evo2 phages (Table 1). We have added this new data into the main text and supplemental figures and tables. Future work will focus on obtaining comprehensive analysis of a wide range of phage families.

(3) Testing additional antibiotics

We also expanded our trade-off analysis include wider range of antibiotic classes (Table S3). Overall, the loss of resistance appears to be confined to b-lactams.

(4) Genetic analysis of ORF141

In order determine the function of ORF141, which is mutated in Evo2, we attempted to clone wild-type ORF141 into a staphylococcal plasmid and perform complementation assays with Evo2. Unfortunately, obtaining the plasmid-borne wild-type ORF141 has proven to be tricky, as all clones developed frameshift or deletions in the open reading frame. We posit that the gene product of ORF141 is toxic to the bacteria. We are currently working on placing the gene under more stringent expression conditions but feel that these efforts fall outside of the scope of this paper.

(5) Testing the effect of single mutants

Our genomic analysis showed that phage-resistant MRSA evolved multiple mutations following phage infection, making it difficult to determine the mechanism of each mutation alone. For example, phage-resistant MW2 and LAC evolved nonsense mutations in transcriptional regulators mgrA, arlR, and sarA. To test whether these mutations alone were sufficient to confer resistance, we obtained MRSA strains with single-gene knockouts of mgrA, arlR, and sarA and tested their ability to resist phage. We observed that deletion of mgrA in the MW2 resulted in a modest reduction in phage sensitivity (Figure S7). However, we did not the observe any changes in the other mutant strains. These results suggest that phage resistance in these strains is likely caused by a combination of mutations. Determining the mechanisms of these mutations is the focus if our future work.

(6) Transcriptomics of phage-resistant MRSA strains

To further assess the effects of the phage resistance mutations, we performed bulk RNA-seq on phage-resistant MW2 and LAC strains and compared their differential expression levels to the respective wild-type strains. We picked these strains because our genomic data showed that they had evolved mutations in known transcriptional regulators (e.g. mgrA). Our analysis shows that both strains significantly modulate their gene expression (Figure 4). Notably, both strains upregulate the cell wall-associated protein ebh, while downregulating several genes involved in quorum sensing, virulence, and secretion. We have included this new data in Figure 4 and Table S5 and added an entire section in the manuscript discussing these results and their implications.

(7) Co-treatment of MRSA with phage and b-lactam

We performed checkerboard experiments on MRSA strains with phage and b-lactam gradients (Figure 6). We found that under most conditions, MRSA cells were only able to recover under low phage and b-lactam concentrations. Notably, these recovered cells were still phage resistant and b-lactam sensitive. However, under one condition where MW2 was treated with FStaph1N and b-lactam, we found that some recovered cells still had high levels of b-lactam resistance, showing a distinct mutational profile. We discuss these results in detail in the main text.

Reviewer # 1:

Strengths:

Phage-mediated re-sensitization to antibiotics has been reported previously but the underlying mutational analyses have not been described. These studies suggest that phages and antibiotics may target similar pathways in bacteria.

We thank Reviewer 1 for this assessment. We hope that the data provided in this work will help stimulate further inquiries into this area and help in the development of better phage-based therapies to combat MRSA.

Weaknesses:

One limitation is the lack of mechanistic investigations linking particular mutations to the phenotypes reported here. This limits the impact of the work.

We acknowledge the limitations of our initial analysis. We note (and cite) that separate studies have already linked mutations in femA, mgrA, arlR, and sarA with reduced b-lactam resistance and virulence phenotypes in MRSA, but not to phage resistance. For the other mutations, we could not find literature linking them to our observed phenotypes. We analyzed the effects of single gene knockouts of mgrA, arlR, and sarA on MRSA’s phage resistance. However, as shown above, the results only showed modest effects on phage resistance in the MW2 strain (see Figure S7 and lines 309-317). We therefore believe that mutations in single genes are not sufficient to cause the trade-offs in phage/ b-lactam resistance. Because each MRSA strain evolved multiple mutations (e.g. MW2 evolved 6 or more mutations), we feel that determining the effects of all possible permutations of those mutations was beyond the scope of the paper.

However, to bridge the mutational data with our phenotypic observations, we performed RNAseq and compared the transcriptomes of un-treated and phage-treated MRSA strains (see Figure 4, Table S5, and lines 337-391). Our results show that phage-treated MRSA strains significantly modulate their transcript levels. Indeed, some of the changes in gene expression can explain for the phenotypic observations (e.g. overexpression of ebh can lead to reduced clumping). Further, the results shown some unexpected patterns, such as the downregulation of quorum sensing genes or genes involved in type VII secretion.

Another limitation of this work is the use of lab strains and a single pair of phages. However, while incorporation of clinical isolates would increase the translational relevance of this work it is unlikely to change the conclusions.

We thank the reviewer for this suggestion. We would like to clarify that MW2, MRSA252, and LAC are pathogenic clinical isolates that were isolated between 1997 and 2000’s. However, we acknowledge that, because these 3 strains have been propagated for many generations, they might have acquired laboratory adaptations. We therefore obtained 30 USA300 clinical strains that were isolated in more recent years (~2008-2011) and tested our phages against them. We note that these clinical isolates (generously provided by Dr. Petra Levin’s lab) were preserved with minimal passaging to reduce the effects of laboratory adaptation. We found that the Evo2 phage was able to elicit oxacillin trade-offs in those strains as well. (see Table S1, Table S7, Fig 2C, and lines 210 – 225)

For the phages, we had to work with phage(s) that could infect all three MRSA strains. That is why in our initial tests, we focused on FStaph1N and Evo2, both members of the Kayvirus genus. Now in our revised work, we extend our analysis to FNM1g6, a member of the Dubowvirus genus, that also infects the LAC strain, but not MW2 and MRSA252. We find that FNM1g6 is unable to drive trade-offs in b-lactam resistance (see lines 229 – 238). Next, we analyzed the effects of SATA8505, also a member of the Kayvirus genus. Here, we observed that SATA8505 can elicit trade-offs in b-lactam resistance (see Figure S5 and lines 238 – 246). These results suggest that not all staphylococcal phages can elicit these trade-offs and call for more comprehensive analyses of different types of phages.

Reviewer #1 (Recommendations for the authors):

Specific questions:

(1) The Evo2 isolate is an evolved version of phage Staph1N with more potent lytic activity. Is this reflected in more pronounced antibiotic sensitivity?

We did not observe that Evo2-treated MRSA cells showed more sensitivity towards b-lactams. However, we did observe that Evo2 was able to elicit these trade-offs at lower multiplicities of infection (MOI) (see lines 173 – 176 and Figure S2). Further, we did observe that Evo2 caused a greater trade-off in virulence phenotypes (hemolysis and cell agglutination) (see lines 416 - 419 lines 433 – 435, and Figure 5)

In our revisions, we also tested Evo2-treated MRSA against a wide range of antibiotics. We did not observe significant changes in MICs against those agents.

(2) Are there mutations in the SCCmec cassette or the MecA gene after selection against ΦStaph1N?

We did not observe any mutations in known resistance genes SCCmec or blaZ. Furthermore, we did not see any differential expression of those genes in our transcriptomic data (see lines 344 and 346).

(3) The authors report that phage ΦNM1γ6 does not induce antibiotic sensitivity changes despite being effective against bacterial strain LAC. Were mutational sequencing studies performed with the resistant isolates that emerged against this strain? Can the authors hypothesize why these did not impact the virulence or resistance of LAC despite effective killing? How does this align with their models for ΦStaph1N?

We thank the reviewer for that insightful question. In our revised manuscript, we found that ΦNM1γ6 elicits a point mutation in the fmhC gene, which is involved in cell wall maintenance (see lines 326 – 335). To our knowledge, this point mutation has not been linked to phage resistance or drug sensitivity MRSA. Notably this mutation was not observed with ΦStaph1N or Evo2. We therefore speculate that ΦNM1γ6 binds to a different receptor molecule on the MRSA cell wall.

(4) If I understand correctly, the authors attribute these effects of phage predation on antibiotic sensitivity and virulence to orthogonal selection pressures. A good test of this model would be to examine the mutations that emerge in antibiotic/phage co-treatment. This should be done.

We thank the reviewer for this suggestion. As described in the summary section above, we performed checkerboard experiments on MRSA strains with phage and b-lactam gradients (see lines 440 – 494 and Figure 6). We found that under most conditions, MRSA cells were only able to recover under low phage and b-lactam concentrations. Notably, these recovered cells were still phage resistant and b-lactam sensitive. However, under one condition where MW2 was treated with FStaph1N and b-lactam, we found that some recovered cells still had high levels of b-lactam resistance and only limited phage resistance, showing a distinct mutational profile (Figure S6). Under these conditions, we think that the selective pressure exerted by FStaph1N is “overcome” by the selective pressure of the high oxacillin concentration, a point that we discuss in the main text.

Reviewer #2 (Public review):

Summary:

The work presented in the manuscript by Tran et al deals with bacterial evolution in the presence of bacteriophage. Here, the authors have taken three methicillin-resistant S. aureus strains that are also resistant to beta-lactams. Eventually, upon being exposed to phage, these strains develop beta-lactam sensitivity. Besides this, the strains also show other changes in their phenotype such as reduced binding to fibrinogen and hemolysis.

Strengths:

The experiments carried out are convincing to suggest such in vitro development of sensitivity to the antibiotics. Authors were also able to "evolve" phage in a similar fashion thus showing enhanced virulence against the bacterium. In the end, authors carry out DNA sequencing of both evolved bacteria and phage and show mutations occurring in various genes. Overall, the experiments that have been carried out are convincing.

We thank Reviewer 2 for their positive comments.

Weaknesses:

Although more experiments are not needed, additional experiments could add more information. For example, the phage gene showing the HTH motif could be reintroduced in the bacterial genome and such a strain can then be assayed with wildtype phage infection to see enhanced virulence as suggested. At least one such experiment proves the discoveries regarding the identification of mutations and their outcome.

We thank the reviewer for this suggestion. We attempted to clone ORF141 into an expression plasmid and perform complementation experiments with Evo2 phage; however, all transformants that were isolated had premature stop-codons and frameshifts in the wild-type ORF141 insert that would disrupt protein function. We therefore think that the gene product of ORF141 might be toxic to the cells. We are currently working on placing the gene under more stringent transcriptional control but feel that these efforts fall outside of the scope of this paper.

Secondly, I also feel that authors looked for beta-lactam sensitivity and they found it. I am sure that if they look for rifampicin resistance in these strains, they will find that too. In this case, I cannot say that the evolution was directed to beta-lactam sensitivity; this is perhaps just one trait that was observed. This is the only weakness I find in the work. Nevertheless, I find the experiments convincing enough; more experiments only add value to the work.

We thank the reviewer for their comments. Because both phages and β-lactams interface with the bacterial cell wall, we posited that phage resistance would reduce resistance in cell wall targeting antibiotics. In our revisions, we have expanded our analysis to include a much wider range of antibiotic classes, including rifampicin, mupirocin, erythromycin, and other cell wall disruptors, such as daptomycin and teicoplanin. We did not observe any significant changes to the MICs of these other antibiotics (see Table S3 and lines 191-199). It therefore appears that the effects of these trade-offs are confined to beta-lactams.

Version 2

Abstract

Abstract

Bacteriophage (phage) therapy is a promising means to combat drug-resistant bacterial pathogens. Infection by phage can select for mutations in bacterial populations that confer resistance against phage infection. However, resistance against phage can yield evolutionary trade-offs of biomedical relevance. Here we report the discovery that infection by certain staphylococcal phages sensitizes different strains of methicillin-resistant Staphylococcus aureus (MRSA) to β-lactams, a class of antibiotics against which MRSA is typically resistant. MRSA cells that survive infection by these phages display significant reductions in minimal inhibitory concentration against different β-lactams compared to uninfected bacteria. Transcriptomic profiling reveals that these evolved MRSA strains possess highly modulated transcriptional profiles, where numerous genes involved in S. aureus virulence were downregulated. Phage-treated MRSA exhibited attenuated virulence phenotypes in the form of reduced hemolysis and clumping. Despite sharing similar phenotypes, whole-sequencing analysis revealed that the different MRSA strains evolved unique genetic profiles during infection. These results suggest complex evolutionary trajectories in MRSA during phage predation and open up new possibilities to reduce drug resistance and virulence in MRSA infections.

Staphylococcus aureus is one of the most notorious and widespread bacterial pathogens, responsible for hundreds of thousands of severe infections worldwide every year. Methicillin-resistant S. aureus (MRSA) poses a particular clinical threat, as MRSA infections increase mortality, morbidity, and hospital stay, as compared to those caused by methicillin-sensitive S. aureus (MSSA).1 Part of MRSA’s notoriety stems from its strong resistance against the β-lactam family of antibiotics, such as penicillins and cephalosporins, which inhibit the activity of transpeptidase enzymes during peptidoglycan synthesis in bacterial cell walls.2 β-lactams are one of the most commonly prescribed drug classes, with many designated as “Critically Important” antimicrobials by the World Health Organization.3 Thus, MRSA infections pose a considerable public health risk as they are notoriously difficult to treat and are widespread in communities and hospital settings. Indeed, in 2019, MRSA alone accounted for more than 100,000 deaths attributable to drug-resistant infections worldwide.4

A chief mediator of β-lactams resistance in MRSA is the SCCmec cassette, a mobile genetic element that carries the resistance gene mecA. MecA encodes for the penicillin-binding protein 2A (PBP2a), a transpeptidase that has a low affinity for β-lactams.5 This lower affinity permits PBP2a to participate in peptidoglycan synthesis even in the presence of β-lactams, ultimately resulting in cell survival. In addition to mecA, numerous MRSA strains also carry β-lactamases, such as BlaZ, that degrade β-lactams, thus further contributing to drug resistance. Together, these mechanisms can severely limit treatment options against MRSA, with current clinical treatment options relying primarily on vancomycin and daptomycin.6 Both vancomycin and daptomycin are last resort antibiotics against MRSA, and a major concern is the increasing resistance of MRSA to these drugs.7,8 Developing solutions to combat MRSA is a major focus in academia and industry.

Due to its drug resistance and clinical burden, S. aureus is a prime candidate for alternative antimicrobial treatments, suchas bacteriophage (phage) therapy. Phages are viruses that infect and kill bacteria, posing one of the greatest existential threats to bacterial communities, with some estimates suggesting that 40% of all bacterial mortality worldwide is caused by phage predation.9 In phage therapy, lytic phages are administered to kill the bacterial pathogen(s) causing an infection. Phages offer certain advantages over traditional antibiotics: they are highly specific to their hosts by reducing off-target killing; they self-amplify and evolve, enabling the rapid generation of new phage variants with improved activities; and they are generally regarded as safe, as toxicity has been reported only in extremely rare cases in animals and patients.10 Indeed, against S. aureus infections, over a dozen promising case studies and clinical trials have been reported in the past decade.11

Despite these advances, routine use of phage therapy is still met with challenges. Chief among these is the inevitable rise of phage resistance, as phage predation exerts a strong selective pressure on bacterial populations. According to one meta-analysis that focused on phage therapy outcomes, resistance against phage evolved in 75 percent of human clinical cases in which the evolution of resistance was monitored.12 Mutations represent a chief pathway by which bacteria evolve resistance against phage. To date, the best characterized phage resistance mutations involve alterations on cell surface receptor molecules that mediate phage attachment. In many bacteria, these receptors are often proteins or sugar moieties, which are recognized by phage proteins.11,13,14 For example, in Escherichia coli, mutations in the cell-wall protein LamB confer resistance against lambda phage infection, while in S. aureus, mutations that modify wall teichoic acid (WTA) have been shown to limit phage infection.14,15 Complicating the picture, studies have revealed that a plethora of additional host mechanisms, including dedicated anti-phage defense systems, can impact the evolution of resistance against phage.16,17 These problems highlight the importance of developing phage treatment strategies that minimize or capitalize on the evolution of phage resistance.18

A unique aspect of phage therapy is the possibility to exploit evolutionary trade-offs to combat resistant pathogens. A genetic trade-off is defined as an evolved trait that confers a fitness advantage against a particular selective pressure at the expense of reduced fitness against an unselected pressure. Across many different species of bacteria, such trade-offs have been shown to occur between phage resistance and antibiotic resistance (Figure 1A). Phages that bind to a virulence factor or mechanism for antibiotic resistance in the target bacteria are predicted to exert a strong selection pressure on the bacteria to mutate or downregulate the phage-binding target. These changes would confer protection against phage infection but could in turn reduce the resistance or virulence in the bacterium. As an example, in P. aeruginosa, infection by the phage OMKO1, which binds to the outer membrane protein M OprM of MexAB- and MexXY-OprM efflux pumps, drives the evolution of mutations in those genes, leading to the re-sensitization of phage-resistant P. aeruginosa mutants to antibiotics.10,19

Infection by bacteriophage ΦStaph1N drives the loss of β-lactam resistance in MRSA.

A. Schematic of the experimental setup. Drug-resistant (AbxR), phage-sensitive (PhageS) bacterial cultures are infected with phage. The population of infected cells is passaged and allowed to recover. The surviving cell population is resistant to phage infection (PhageR) but has evolved sensitivity to antibiotics (AbxS). B. ΦStaph1N infects MRSA strains MRSA252, MW2, and LAC (left panel). Following infection with ΦStaph1N, evolved cultures of the three MRSA strains are resistant to ΦStaph1N (right panel). C. ΦStaph1N-treated, evolved MRSA strains show significant loss of resistance against oxacillin, compared to the parental strains. Loss of resistance is indicated by the area of bacterial clearance surrounding the antibiotic resistance strip. D. ΦStaph1N treatment causes loss resistance against different β-lactams. Plotted are the fold reductions of minimal inhibitory concentration (MIC) between treated and mock-treated cells. OXI = oxacillin; CEF = cefazolin; AMX = amoxicillin; AMX+CA = amoxicillin & clavulanic acid; VANC = vancomycin.

Little is known about how phage resistance can mediate genetic trade-offs in MRSA. Previous work has shown that phage resistance in S. aureus can proceed through genetic mutations that are not directly involved in phage binding. For example, studies by Berryhill and colleagues demonstrated that phage infection of S. aureus Newman, an MSSA strain, can select for mutations in femA, which is a cytoplasmic enzyme that catalyzes the formation of the pentaglycine bridge of peptidoglycans in S. aureus.20 Rather than serving as a direct phage receptor molecule, femA maintains the integrity of the cell wall, which in turn could be vital for WTA maturation and phage attachment.13,21 Interestingly, a consequence of these femA mutations is increased sensitivity against antibiotics. We therefore asked how phage resistance might impact the physiology of drug-resistant MRSA, with the hopes of identifying genetic trade-offs of potential biomedical relevance.

In the work reported here, we show that infection by staphylococcal phages causes MRSA strains to evolve sensitivity to different types of β-lactam antibiotics and attenuate virulence phenotypes. We found that this loss of resistance and virulence is associated with distinct mutational profiles distinct in each MRSA strain, and that phage-treated, evolved MRSA populations display significant transcriptome remodeling. Unexpectedly, we also discovered a mutant phage with higher activity and a broader host range against MRSA. Findings from our work can help in the development of phage therapies that reduce drug resistance and virulence in pathogenic bacteria.

Results

Identification of ΦStaph1N with activity against multiple MRSA strains

For our studies, we focused on three MRSA strains – MRSA252 (USA200), MW2 (USA400), and LAC (USA300). All three MRSA strains are pathogenic isolates implicated in human disease and are used as representative examples for studying MRSA.22–24 To test for phage susceptibility, we performed plaquing assays with a panel of staphylococcal phages (Table S1, Figure S1). Of the phages tested, one phage, ΦStaph1N, which belongs to the Kayvirus genus, formed plaques on all three MRSA strains (Figure 1B).21,25 Yet, despite its ability to infect all three strains, ΦStaph1N infection was unable to eradicate MRSA cultures. Both MW2 and LAC displayed incomplete lysis in liquid culture at multiplicities of infection (MOIs) of 0.1 or lower (Figure S2). Furthermore, infected cultures of all three MRSA strains could recover back to high cell density after passaging one percent of the culture into fresh media following 24 hours of initial infection. These results suggest that infection by ΦStaph1N selects for resistant mutants that could sweep the population. Indeed, ΦStaph1N was unable to form plaques on recovered MRSA cultures that survived in the initial ΦStaph1N infection (Figure 1B).

Resistance against ΦStaph1N infection sensitizes MRSA against β-lactams

Because both phages and β-lactams interface with the bacterial cell wall, we hypothesized that resistance against ΦStaph1N infection could cause a trade-off in β-lactam resistance in MRSA even in the presence of PBP2a and BlaZ. We first tested the β-lactam sensitivity of the parental MRSA252, MW2, and LAC strains. As expected, all three strains displayed high minimal inhibitory concentrations (MICs) of ≥48 µg/mL against the β-lactams oxacillin (OXA), cefazolin (CEF), amoxicillin (AMX), and amoxicillin & clavulanic acid (AMX+CA), visually indicated by their ability to form lawns surrounding antibiotic strips (Figure 1C, Table S2). The strains were sensitive to vancomycin (VANC; MICs = 1.5 µg/mL), which inhibits cell wall synthesis through a different mechanism than β-lactams. Strikingly, phage-resistant MRSA that survived ΦStaph1N infection displayed a strong reduction in resistance against OXA, CEF, and AMX+CA, with fold reductions in MIC between 10 and 1000-fold (Figure 1D); no change in MIC was observed with VANC or with AMX alone. These results show at a phenotypic level that ΦStaph1N-resistant MRSA loses resistance towards most β-lactams.

We next asked whether this loss of β-lactam resistance depended on the MOI of ΦStaph1N. We infected the three MRSA strains with ΦStaph1N at MOIs ranging from 10−2 to 10−5, isolated the surviving MRSA cells, and tested their MIC against oxacillin (Table S2). For MRSA252, we still observed a ∼3-order of magnitude fold reduction of MIC at an MOI of 10−5. With MW2, the reduction of MIC was markedly decreased with lower phage levels, showing no significant loss at MOIs of 10−3 or lower. For LAC, two replicates displayed a reduction of MIC by an order of magnitude at an MOI of 10−4, while the third replicate did not display any change. These results show that for MRSA252, ΦStaph1N MOIs as low as 10−5 can still drive the loss of resistance, while for MW2 and LAC, higher MOIs of phage are needed to ensure the same outcome of reduced β-lactam resistance.

Discovery of a mutant ΦStaph1N with enhanced activity against MRSA

For ΦStaph1N, we noticed that while the phage could plaque on all three MRSA strains, its plaque-forming efficiency was reduced on the MW2 and LAC strains (Figure 1B, Figure S1). ΦStaph1N plaques on MW2 and LAC bacterial lawns were hazy, and the overall efficiency of plaquing was approximately two orders of magnitude less than that on MRSA252. Unexpectedly, we consistently observed smaller, clear plaques arising in the larger, hazy plaques of LAC (Figure S3A); notably this did not appear in MW2. We hypothesized that these clear plaques were caused by a mutant form of ΦStaph1N that evolved higher lytic activity. We isolated phage clones from these single plaques and tested their activity against MRSA. This mutant phage, which we called Evo2, plaques on LAC and MW2 strains with higher efficiency, displaying comparable plaquing to MRSA252 (Figure 2A). In growth experiments, we further observed that Evo2 lyses MRSA cultures at lower MOIs compared to ΦStaph1N (Figure S2). Evo2 exhibits lytic activity against MW2 and LAC even at an MOI of 10−4, a concentration at which ΦStaph1N does not show any detectable activity against the two strains.

Evo2 is a variant of ΦStaph1N with higher activity against MRSA.

A. Evo2 shows comparable infectivity towards MRSA252 but improved infectivity towards MW2 and LAC, relative to ΦStaph1N. The same plaquing data is also shown in Figure S3A. B. Similar to ΦStaph1N, Evo2 infection reduces β-lactam resistance in MRSA. C. Evo2 infection reduces the MIC against oxacillin in clinical isolates of USA300 (ADLs).

We sequenced the genome of Evo2 to determine the genetic mechanism driving this enhanced activity. We observed a single point mutation in ORF141 that induces a premature stop codon (Figure S3B). Sequence analysis with HHpred predicts ORF141 to be a putative DNA binding protein with an HTH motif (PDB: 2LVS, E-value: 2.5e-9). We speculate that this protein is a transcriptional regulator that when inactivated by a nonsense mutation, increases ΦStaph1N infectivity. Future studies will center on determining the mechanism of this mutation and why Evo2 only evolved in the LAC strain.

Given Evo2’s enhanced activity against MRSA, we asked how predation by Evo2 affected β-lactam resistance. We infected MRSA252, MW2, and LAC with Evo2 at an MOI of 0.1 and measured the MICs against β-lactams after 48 hours of passaging. Similar to ΦStaph1N, infection by Evo2 reduced the MICs of the three MRSA strains against OXA, CEF, and AMX+CA, while MICs against AMX alone and VAN did not change significantly (Figure 2B). Expanding on the different classes of antibiotics, we tested whether Evo2 predation could impact the susceptibility to the transcription inhibitor rifampicin; the translation inhibitors erythromycin and mupirocin; and the cell envelope disruptors teicoplanin, fosfomycin, and daptomycin (Table S3). We found that the MICs of these antibiotics did not change significantly, with a few exceptions: in some cases, Evo2-resistant LAC became sensitized to fosfomycin and daptomycin; furthermore, one replicate of Evo2-resistant MRSA252 evolved sensitivity to teicoplanin. However, overall the MIC reduction in these cases was not as dramatic as the MIC reduction seen against β-lactams.

Finally, we examined how different MOIs of Evo2 impacted β-lactam resistance in MRSA (Table S2). We infected the three MRSA strains with Evo2 at varying MOIs from 10−2 to 10−5 and measured the MIC against oxacillin of the evolved MRSA. Across all three strains, we found that replicate cultures across different MOIs were unable to recover growth following Evo2 infection (Table S2). However, cultures of MRSA252, MW2 and LAC that did regrow displayed a loss of oxacillin resistance, between 10- to 1000-fold. Thus, overall Evo2 displayed a higher infectivity against the MRSA and a greater potency in reducing β-lactam resistance.

Evo2 is broadly active against recent clinical isolates of S. aureus USA300

MRSA252, MW2, and LAC were isolated in 1997, 1998, and in the early 2000s, respectively. We therefore tested if Evo2 can infect more recent S. aureus clinical isolates. We compared the plaquing efficiency of Evo2 and ΦStaph1N against 30 USA300 strains that were isolated between 2008 and 2011 at St. Louis Children’s Hospital (Table S1).26 We observe dramatic variation in the plaquing efficiency of ΦStaph1N and the 30 strains, while Evo2 exhibited a higher plaquing efficiency in the majority of the 30 strains (Table S4). We then tested how infection of Evo2 impacted OXA resistance in 12 (ADL1-12) of these clinical isolates. We infected the strains with Evo2 at an MOI of 0.1 and, if a phage-resistant population was recovered, measured the OXA MIC after 48 hours of passaging. Interestingly, after 15 independent challenges with Evo2, we were not able to recover phage-resistant populations from ADL1, 5, 6, and 12. This suggests that Evo2 resistance acquisition is a rare event in these strains. In the rest of the strains, we observed that similarly to MRSA252, MW2, and LAC, the OXA MIC was reduced between 10-to 100-fold after Evo2 infection (Figure 2C). Overall, these results highlight the broader host range and activity of Evo2.

Effect on β-lactam MIC by other phages

We asked whether two additional phages from our collection could elicit MIC reduction against oxacillin in MRSA. The MRSA LAC strain is sensitive to infection by ϕNM1γ6, a lytic version of the temperate phage ΦNM1 of the Dubowvirus genus, derived from the S. aureus Newman strain.27–29. In LAC, ΦNM1γ6 displays plaquing comparable to that of ΦStaph1N, while also showing activity against some of the clinical USA300 isolates (Figure S1, Table S4). Therefore, we infected LAC with ΦNM1γ6 at an MOI of 0.1 and measured the MIC against β-lactams of the surviving cells. While the recovered LAC cultures exhibited resistance against ΦNM1γ6, they did not show a reduction in MIC (Figure S4), suggesting that phage resistance caused by ΦNM1γ6 is uncoupled from β-lactam resistance. We also isolated from the environment a second phage of the Kayvirus genus, called SATA8505,30 and tested its activity against MRSA. SATA8505 is active against MRSA252, MW2, and LAC (Figure S5), and infection of the three strains caused a rise of phage resistance in MRSA252, MW2, and LAC; (Figure S5A). Similar to ΦStaph1N and Evo2, cells resistant to SATA8505 showed a strong loss of oxacillin resistance (Figure S5B). Altogether, our results with these two phages suggest that the ability to reduce β-lactam resistance is not a universal feature across all staphylococcal phages and call for a more comprehensive analysis of staphylococcal phages and their ability to elicit β-lactam trade-offs.

Genomic mutations in MRSA strains following phage infection

Following our phenotypic analyses, we examined the genomes of the phage-resistant MRSA. We first sequenced the genomes of three clonal isolates (A-C) from each MRSA strain that underwent ΦStaph1N, Evo2, or a mock infection. We observed that each MRSA strain evolved distinct mutation profiles (Figure 3A). Irrespective of the strain and phage treatment, most mutations were predicted to be substitutions, followed by truncations (Figure S6). Cluster of Orthologous Genes (COG) variants associated with transcription, cell wall/membrane/envelope biogenesis, coenzyme transport and metabolism, and defense mechanisms were the most commonly found categories. Mutations in annotated genes that appeared at least twice across the clonal replicates are summarized in Table 1. Information on all detected genetic variants is listed in Supplementary Table 1.

Phage infection of MRSA strains produces distinct mutational profiles.

A. Coding sequences (CDS) with mutations from the three MRSA strains following phage treatment or mock treatment. For each strain, three isolates were sequenced and their mutations identified. Mutations are color-coded based on the number of occurrences among the three replicates. B. Categories of genes with mutations that arose in each MRSA strain and treatment condition.

One plausible hypothesis explaining the loss of β-lactam resistance is that phage infection selected for a defective SCCmec or blaZ. However, we did not observe any mutations in the two loci. Instead, all MRSA strains exhibited mutations in ancillary genes implicated on the loss in β-lactam resistance. In MRSA252, both ΦStaph1N and Evo2 infection were selected for frameshift or nonsense mutations in the femA gene that would inactive the protein product. As discussed above, femA is required for the synthesis of the pentaglycine branch on S. aureus Lipid II, the peptidoglycan precursor (Table 1). Deletions of femA have been shown to increase susceptibility to β-lactams even when PBP2a (encoded by mecA) is expressed, thus providing a genetic mechanism for how some MRSA252 cells lose β-lactam resistance after phage selection.43,44 At the same time, we found the presence of other uncharacterized mutations in phage-resistant MRSA252. For example, clone A of ΦStaph1N-treated cells carried 2 mutations: a frameshift in femA and a substitution mutation in an uncharacterized protein; meanwhile, clone B displayed a substitution mutation in pfkA, a predicted ATP-dependent 6-phosphofructokinase and mutation in an intergenic region; clone C showed a substitution mutation in a putative transport protein, called yueF (Figure 3B, Supplementary Table 1). The role of these mutations in mediating phage resistance or β-lactam sensitivity, if any, remains unknown.

In MW2, we found mutations in two transcriptional regulators, mgrA and sarA (Figure 3A, Table 1). Both mgrA and sarA belong to the family of MarR (multiple antibiotic resistance regulator)/SarA (staphylococcal accessory regulator A) proteins, which regulate drug resistance and virulence in S. aureus.31–33 In ΦStaph1N-treated MW2, only mgrA was mutated, while in Evo2-treated MW2, clones also showed nonsense mutations in sarA. We also found mutations in metK and ytqA which are both predicted to be associated with S-adenosylmethione (SAM): metK synthesizes SAM, while ytqA belongs to the radical SAM enzyme family and is predicted to be involved in tRNA modification.41,47 Phage-treated MW2 also displayed mutations in other genes, including tcaB and murE. Notably, each clonal replicate had multiple mutations in the genome, while by contrast untreated MW2 cells only displayed a deletion in an intergenic region that is not present in any of the phage-treated samples. These findings suggest that MW2 could be amassing multiple mutations during the course of phage infection.

For LAC, we observed a third, distinct mutational pattern (Figure 3A). Of note, we found nonsense mutations in arlR, which is part of the arlRS two-component signaling system (Table 1). The activity of arlRS has been implicated in S. aureus virulence, pathogenicity, and oxacillin resistance.36,37 Further, we observed substitution mutations in spoVG, which is a transcription factor regulating the expression genes involved in a variety of functions, including cell wall metabolism (Table 1).38 Indeed, spoVG activates the expression of femA.39 Studies have shown that spoVG modulates β-lactam antibiotic resistance by modulating cell wall synthesis.39 Similar to the other two MRSA strains, phage-treated clones of LAC showed multiple mutations in their genomes. We observed mutations in prsA and bioA that appeared in both the mock and phage treatment conditions, suggesting that these mutations do not arise due to phage selection.

Altogether, our results show that phage-infected MRSA strains acquire distinct mutational profiles. These mutations likely work in concert to promote phage resistance and β-lactam sensitivity, making it challenging to determine the mechanistic contributions of individual mutations. For example, we observed that the genes mgrA, sarR, and arlR evolved nonsense mutations, which would result in truncated, potentially non-functional protein products. We therefore tested if single knockout mutants of these genes alone are sufficient to confer resistance to ΦStaph1N and Evo2 (Figure S7). In MW2, the mgrA knockout resulted in a modest reduction in plaquing of Evo2 and ΦStaph1N. However, none of the remaining mutants in either the MW2 or LAC background conferred resistance. Prior experimental studies have also shown that phage resistance in S. aureus can arise from the disruption of single genes directly involved in the synthesis and modification of WTA, such as tagO.51 Some MRSA strains also alter cell wall glycosylation through dedicated genes encoded on prophages.14 However, we did not see any mutations in genes directly involved in WTA synthesis. Our results thus highlight how MRSA can take on unique mutational pathways under phage selection.

Finally, we examined the mutational profile in ΦNM1γ6-resistant LAC populations Because infection with ΦNM1γ6 did not result in a decrease in OXA resistance, we hypothesized that mutations that arose in LAC following ΦNM1γ6 would be distinct from those following Evo2 infection. We found that the two genes were mutated across three clonal isolates from different resistant populations: bioA and fmhC. As seen previously, mutations in bioA appeared in the mock treatment, suggesting that that the mutations arose independently of phage selection. On the other hand, LAC showed a missense mutation in fmhC (H21D, Table 1, Supplementary Table 1). FmhC and its homologue fmhA pair with femA and femB to incorporate Gly-Ser dipeptides into peptidoglycan cross-bridges.50 However, the mechanism of the H21D mutation is unknown, and to our knowledge, mutations in fmhC have not been associated with phages resistance in S. aureus before.

Phage-treated MRSA strains display broad changes to their transcriptomes

Our genomic analysis revealed that phage-treated MRSA evolved mutations in a variety of transcriptional regulators, some of which are known to affect MRSA virulence. We therefore hypothesized that the mutations in these regulators would fundamentally alter the transcriptional profile of the treated MRSA. To test this, we performed bulk RNA-seq experiments on MW2 and LAC strains that were treated with the phage Evo2 and compared their transcription profiles to those of untreated strains (Figure 4, Table S5). We observed significant changes in gene expression in both MW2 and LAC. Notably, mirroring the trend seen in the mutational data, we did not observe significant changes in the expression of genes in the SCCmec cassette or blaZ present in both MW2 and LAC.

Phage infection changes the transcriptomic profile of MRSA.

Differential expression analysis was performed on the transcriptomes of MW2 (top panel) and LAC (bottom panel). For both strains, Evo2-infected samples were compared to uninfected controls. 3 biological replicates were analyzed for each condition. Horizontal dotted lines represent an adjusted p-value cut-off of 0.002, while vertical dotted lines represent a log2 fold change of −2 or 2 in expression. Transcripts with a log2 fold change between −2 or 2 and a p-value >0.002 are labeled as grey dots (Not significant, NS); transcripts that pass either the fold change or p-value cutoff (but not the other) are represented as blue and green dots, respectively; transcripts that pass both cutoffs are shown as red dots. Genes discussed in the main text are labeled. Data for all the transcripts with significant fold changes is shown in Table S5.

We first compared our expression data against transcriptomic studies from previous studies. For example, Horswill and colleagues have shown that deletions of arlRS and mgrA de-represses the extracellular matrix binding protein ebh, resulting in significantly higher expression levels of the gene. In addition, these deletions also increased the expression of urease genes involved in the urea TCA cycle.33,37 In our experiments, both MW2 and LAC strains evolved nonsense mutations in mgrA and arlR, respectively. We therefore hypothesized that these mutations would mimic the effects of gene deletions and likewise yield elevated transcript levels of ebh and urease genes. Aligning with our hypothesis, differential expression data from both MW2 and LAC displayed a log2 fold changes of >7 for ebh and >3 for ureABCDEFG genes (Table S5).

Additionally, both MW2 and LAC strongly upregulated several genes involved in cell wall maintenance. These include lytN, which is a murein hydrolase involved in the cross-wall compartment of S. aureus, and fmhC, which, as described previously, incorporate Gly-Ser dipeptides into pentaglycine cross-bridges in the S. aureus peptidoglycan cell wall. Overexpression of lytN has been shown to damage the cell wall, which in turn is alleviated by overexpression of fmhC.50 Interestingly, fmhC overexpression is linked to increased β-lactam sensitivity and thus may contribute to the loss of β-lactam resistance phenotypes we observed in the MRSA strains.

MW2 and LAC also downregulated numerous genes, many of which are known virulence factors. Both strains reduced transcript levels of genes in the locus of the type VII secretion system (ess locus), staphyloxanthin biosynthesis (crtM, crtN, crtP), and quorum sensing (e.g. agrA) (Figure 4, Table S5). Individually, these pathways have been shown to bolster the ability of S. aureus to establish infection and evade the host immune system.52–54 Our results suggest that infection by Evo2 can lead MRSA to reduce the expression of all of these pathways, which could reduce the virulence of S. aureus.

Finally, we noted that both MW2 and LAC showed transcriptional changes that appear to be strain specific. For example, MW2 saw significant increase in the virulence factor spa, known to interfere with the host immune response and interface with other bacterial species. The presence of cell wall-bound Protein A has also been shown to decrease phage absorption, likely by masking WTA.55 Further in MW2, we found that tarM, which adds α1,4-GlcNAc to WTA, was strongly up-regulated (log2 ratio = 6.63). This is in line with previous findings showing that elevated tarM and α1,4-GlcNAc-WTA can lead to phage resistance in MRSA. The LAC showed an increase in the expression (log2 fold change >3.8) of the hemolytic cytotoxin genes lukD/E, which lyses host cells and targets neutrophils.56 We do not know whether the increased expression of these genes results in a greater level of protein production and secretion, but these transcriptional changes could represent a potential “trade-up” associated with phage resistance. Additional studies will be needed to fully assess the physiological and ecological effects of these upregulated genes in MRSA. Altogether, our RNA-seq results suggest that phage infection and resistance in MRSA cause significant transcriptional changes across a wide range virulence, metabolic, and cell-wall associated genes.

Phage-treated MRSA strains display reduced virulence phenotypes

S. aureus is a highly virulent pathogen, relying on a vast array of toxins and immune evasion proteins to promote infection.57 MW2 and LAC strains as models of MRSA virulence.33,34 In light of our mutational and transcriptomic data, we hypothesized that phage-treated MRSA cells would display altered virulence phenotypes, in addition to reduced β-lactam resistance. We first tested the ability of MRSA strains that survived phage predation for their ability to form biofilms in a Crystal Violet assay. We found that Evo2 infection of MRSA252 resulted in a significant reduction in Crystal Violet absorption compared to the parental strain. However, we show no significant difference in Crystal Violet absorption between parental, mock- and phage-treated MW2 and LAC strains (Figure S8).

Next, we tested whether phage infection could affect the hemolysis of rabbit blood cells. Hemolysis is mediated by the secretion of toxins, notably alpha toxin encoded by the gene hla, and plays an important role in MRSA infection.58 Expression of these toxins is regulated by virulence pathways that comprise numerous transcription factors, including mgrA, arlR, and sarA. Furthermore, in our RNA-seq results, we found that phage-resistant MRSA strains showed reduced expression of other cytotoxins. Parental MW2 and LAC colonies lysed rabbit blood cells on blood agar plates, producing distinct halos of clearance around the bacterial cells. For untreated LAC, the total area of hemolysis was on average 3-fold larger than that of untreated MW2 (∼210 mm2 vs ∼70 mm2, respectively); with MRSA252, by contrast, lysis was not detected (Figure 5A). Following treatment with ΦStaph1N, we observed that MW2 and LAC displayed a reduced area of hemolysis by 4 to 5-fold. With Evo2-treated cells, we found that in MW2 the fold reduction was comparable to that of ΦStaph1N-treated cells. However, for Evo2-treated LAC, loss of hemolysis was even more pronounced, with two of the replicates showing no detectable hemolysis. We note that neither MW2 nor LAC showed a reduction in transcript expression of hla. We posit that the loss of hemolysis could be driven by an inability of phage-evolved MRSA to secrete the toxin.

Phage treatment of MRSA results in attenuated virulence phenotypes.

A. MW2 and LAC strains display hemolytic activity on rabbit blood agar plates, while MRSA252 does not. B. Phage-treated MW2 and LAC strains display reduced hemolysis compared to uninfected cells. C. Surviving cultures of MW2 and LAC treated with either ΦStaph1N (blue) or Evo2 (red) show reduced clumping rates compared to mock untreated cells (teal).

We next tested how phage infection affected cell agglutination (or clumping) in MRSA. S. aureus binds to fibrinogen, forming protective aggregates of bacterial cells. Clumping is thought to have several functions in the context of staphylococcal infections, facilitating adhesion to the host tissue. Clumps are also likely to be more resistant to clearance by the immune system, partly because they may be too large to be phagocytosed by neutrophils.34 In our transcriptional data, we noted that several cell surface proteins known to reduce cell clumping, such as ebh, were over-expressed in phage-resistant MRSA. Horswill and colleagues found that de-repression of ebh reduces clumping. Indeed, phage-treated MW2 and LAC displayed less clumping than the mock-treated or parental strain. For MW2, ΦStaph1N infection resulted in a modest reduction, while Evo2 infection resulted in a reduction of approximately 3-fold (Figure 5B). In LAC, we found that both ΦStaph1N and Evo2 treatment resulted in comparable reductions of clumping in surviving cells. Overall, these phenotypic results align with our genetic and transcriptomic data, showing that phage infection can drive MRSA populations to reduced virulence phenotypes.

Combination treatment between Evo2 and β-lactam

The aforementioned results suggest that MRSA cells evolve phage resistance following infection, which is associated with trade-offs in virulence and β-lactam resistance. We next asked how MRSA populations would evolve under co-treatment with phage and β-lactam. In principle, these two simultaneous selective pressures could drive the evolution of resistance against both the phage and antibiotic, negating the trade-offs in drug resistance. To test this, we performed checkerboard assays with phage and oxacillin on MRSA252, MW2, and LAC. Serial dilutions of Evo2 or ΦStaph1N were mixed with serial dilutions of oxacillin on a 96-well plate (Figure 6A), after which MRSA strains were added to the plate and allowed to grow for 24 hours. Following 24 hours, one percent of the culture in each well was transferred into a fresh plate well with nonselective media, and the cultures were allowed to grow for another 24 hours (48 hours total). Throughout the experiment, the cell density was monitored by measuring the optical density in each plate well.

Co-treatment of MRSA with bacteriophage and β-lactam.

A. Checkerboard assays of MRSA strains with gradients of oxacillin and Evo2 (top panels) or ΦStaph1N (bottom panels). The oxacillin gradient is a 2-fold serial dilution of drug concentration (µg/mL), while the phage MOI gradient is a 10-fold serial dilution of MOI. The rows and columns of each plate are labeled with letters and numbers, respectively. The black-white gradient in each well reflects the optical density of the culture and is the mean value from three biological replicates. MRSA strains co-treated with oxacillin and (B) Evo2 or (C) ΦStaph1N were tested for their phage resistance and oxacillin resistance. The letter/number combination reflects the well from which the cells were picked for analysis. Wells that could not produce a viable culture are labeled as NG (no growth). For wells that regrew, we calculated the efficiency of plaquing (EOP) of phage and measured the fold reduction in oxacillin MIC. Cultures that showed no detectable viral plaques are labeled as resistant (R).

We first examined how MRSA grew in combinations of Evo2 and antibiotic (Figure 6A, top row). For MRSA252, cells grew at low levels of phage (MOI of 0.01 or less) and oxacillin (<0.125 µg/mL) (Figure 6A). LAC displayed greater sensitivity, displaying no detectable growth after 48 hours in the presence of phage, irrespective of the presence of oxacillin. For MW2, cells showed limited growth at MOIs <1 and oxacillin levels <0.125 µg/mL. For each strain, we picked cells from wells with an OD600 >0.5 and inoculated them into fresh, non-selective media. As a control, we also regrew cells that were treated with neither phage nor oxacillin (well B1). For MRSA252, two (E2 and F2) wells contained viable MRSA cells, forming turbid cultures; for MW2, six (D4, E2, E3, E5, F2, and F3) wells picked regrew (Figure 6B). We posit that cells in the failed cultures had reduced viability from the phage and antibiotic treatment. We tested these regrown cultures for their phage and oxacillin susceptibility. As expected, MRSA252 and MW2 cells from the B1 control wells were sensitive to Evo2, but resistant to oxacillin. In MRSA252, cells from E2 and F2 were resistant to Evo2 infection and exhibited a 1000-fold decrease in MIC against oxacillin. Similarly, in MW2, the six viable cultures exhibited Evo2 resistance and a 10- to 100-fold decrease in MIC. Altogether, these results match with those of MRSA that underwent single selection with Evo2.

We next analyzed how MRSA grew under ΦStaph1N/β-lactam combinations (Figure 6A, bottom row). Neither MRSA252 nor LAC could grow in any ΦStaph1N/oxacillin combination; MW2, by contrast, grew across under a wide range of ΦStaph1N/oxacillin combinations. When ΦStaph1N-infected MW2 was treated with high (16 µg/mL) levels of oxacillin, recovered cells (wells D11 and E11) showed no reduction in MIC against oxacillin (Figure 6C). However, the plaquing efficiency of ΦStaph1N was reduced by 3 to 6 orders of magnitude. Whole genome sequencing revealed that these cells evolved a unique set of mutations different from those seen in the single phage treatment conditions (Table S6). However, when ΦStaph1N-infected MW2 was co-treated with low levels (<0.125 µg/mL) of oxacillin, recovered cells (C3) displayed strong phage resistance and a 100-fold reduction in oxacillin resistance, mirroring phenotypes observed in the single phage treatment experiments.

We note that of the three MRSA strains tested, MW2 is the least sensitive to ΦStaph1N infection. We posit that the selective pressure exerted by high levels of β-lactam dominates over the selective pressure of ΦStaph1N, leading to the evolution of cells with continued β-lactam resistance and limited phage resistance. However, at lower β-lactam levels, the pressure exerted by phage infection predominates, leading to the rise of cells with complete phage resistance and trade-off in β-lactam resistance. This dose-dependent selection by oxacillin is not observed when a more active phage (e.g. Evo2) is used. While limited to one MRSA strain, these results suggest that different phage/β-lactam combinations can produce divergent evolutionary outcomes in MRSA, each with potential clinical implications.

Discussion

Here we report the discovery that infection by certain phages can drive MRSA populations to evolve favorable genetic trade-offs between phage and β-lactam resistance. Exploiting genetic trade-offs has been proposed as a means to combat resistance in bacterial pathogens.18 Not only could phage treatments reduce the bacterial load of an infection, but also potentially resensitize bacterial populations to antibiotics against which they were previously resistant. We show that MRSA strains infected with Kayviruses ΦStaph1N, Evo2, and SATA8505 evolved resistance against phage, yet developed up to a 1000-fold loss in their MICs against β-lactam antibiotics. In addition, these evolved MRSA display reduced virulence phenotypes such as lower levels of hemolysis and clumping. Our findings show that phages can resensitize MRSA to β-lactams and even decrease their virulence, which are outcomes of significant biomedical value. However, not all staphylococcal phages can mediate these trade-offs: infection with ΦNM1γ6, a phage of the Dubowvirus genus, did generate phage resistance in MRSA, but did not produce a drop in β-lactam resistance. A major future direction will be to determine which types of phages elicit these beneficial evolutionary trade-offs in MRSA.

Our results also paint a complex picture of MRSA evolution during phage infection. Whole genome sequencing revealed that MRSA strains evolved distinct mutation profiles following phage infection, suggesting a multitude of evolutionary paths that different bacterial strains can undertake to evolve resistance against phage. Not only did MRSA strains evolve distinct mutations, but individual, phage-resistant clones accumulated multiple mutations in their genome. Nonetheless, different MRSA displayed a convergence of phenotypes in the form of phage resistance, reduced β-lactam resistance, and attenuated virulence. We posit that this convergence is caused by the involvement of the cell wall in all three phenotypic outcomes. Phages must interface with the cell wall, β-lactams target proteins associated with cell wall maintenance, and many S. aureus virulence factors are embedded within the cell wall or are secreted through it. Thus, any modifications to the integrity or chemical composition of the cell wall by phage resistance will impact β-lactam sensitivity and virulence. Cell wall maintenance is controlled by numerous genes, ranging from single proteins involved in cell wall synthesis, such as femA, to transcriptional regulators, such as mgrA, that control the expression of cell wall synthesis genes. Thus, the mutational patterns observed in each MRSA strain could reflect genetic solutions that enable the bacteria to adapt to the phage predation, while also maximizing the fitness for that particular strain.

Strikingly, MRSA heavily modulated transcriptional profiles following phage infection. We believe these altered expression profiles are a consequence of the genomic mutations that emerged in the various transcriptional regulators. We observed that evolved cells downregulated genes involved in quorum sensing, type VII secretion, and a variety of toxins. It is intriguing to speculate how the down-regulation of these genes impacts MRSA interactions with other bacteria occupying the same ecological niche and with the host immune system. At the same time, MRSA strains also upregulated expression of select virulence factors, such as spA, which could represent “trade-ups.” Trade-ups are thought of as non-selected traits that are enhanced following selection (e.g. cross-resistance between phage and antibiotic), which from a therapeutic perspective might be undesirable.59 Future work will focus on assessing the risk of these trade-ups in light of the clinical benefit of reduced resistance and virulence.

Drug resistance in bacterial pathogens is an evolutionary problem and will require evolution-guided solutions to mitigate. Our findings highlight the ability of phages to dramatically alter the evolution and physiology of drug-resistant MRSA. Select phage treatments can force bacterial populations down evolutionary paths that make them vulnerable to antibiotics or the host immune system. Critically, this permits the re-deployment of agents that would otherwise remain ineffective, buying time for new drug discoveries. We therefore hope that our work may suggest avenues of research into new phage-based treatment strategies against MRSA and other drug resistance pathogens.